---
title: "Project 1"
author: "Naycari de Luna and Marc Robert Grabiel"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    toc_depth: 3
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
---
```{r setup, include = FALSE}

#### Load necessary packages ####
# * These packages are not necessary to complete the assignment and or only used 
#   to provide an example. 
packages <- c("knitr", "magrittr", "geosphere", "RCurl", "dplyr", "kableExtra")

install_me <- packages[!(packages %in% installed.packages()[, "Package"])]
if (length(install_me)) install.packages(install_me)

library(knitr)
library(magrittr)
library(geosphere)
library(RCurl)
library(dplyr)
library(kableExtra)
```

## Background
The World Health Organization has recently employed a new data science initiative, *CSIT-165*, that uses data science to characterize pandemic diseases. 
*CSIT-165* disseminates data driven analyses to global decision makers.

*CSIT-165* is a conglomerate comprised of two fabricated entities: *Global Health Union (GHU)* and *Private Diagnostic Laboratories (PDL)*. 
Your and your partner's role is to play a data scientist from one of these two entities.

## Data
> [2019 Novel Coronavirus COVID-19 (2019-nCoV) Data Repository by John Hopkins CSSE](https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series)

Data for 2019 Novel Coronavirus is operated by the John Hopkins University Center for Systems Science and Engineering (JHU CSSE).
Data includes daily time series CSV summary tables, including confirmations, recoveries, and deaths. 
Country/region are countries/regions hat conform to World Health Organization (WHO).
Lat and Long refer to coordinates references for the user. 
Date fields are stored in MM/DD/YYYY format.

```{r echo=FALSE}

confirmed_download <- getURL("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv")
confirmed <- read.csv(text=confirmed_download, stringsAsFactors = FALSE)

recovered_download <- getURL("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_recovered_global.csv")
recovered <- read.csv(text=recovered_download, stringsAsFactors = FALSE)

deaths_download <- getURL("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv")

deaths <- read.csv(text=deaths_download, stringsAsFactors = FALSE)


```

## Project Objectives

### Objective 1
```{r ob1}

confirmed_origin <- confirmed[which.max(confirmed$X1.22.20), c("Province.State", "Country.Region")]
confirmed_origin

confirmed_country <- as.character(confirmed_origin[[c("Country.Region")]])
confirmed_state <- as.character(confirmed_origin[[c("Province.State")]])

deaths_origin <- deaths[which.max(deaths$X1.22.20), c("Province.State", "Country.Region")]
deaths_origin

deaths_country <- as.character(deaths_origin[[c("Country.Region")]])
deaths_state <- as.character(deaths_origin[[c("Province.State")]])

recovered_origin <- recovered[which.max(recovered$X1.22.20), c("Province.State", "Country.Region")]
recovered_origin

recovered_country <- as.character(recovered_origin[[c("Country.Region")]])
recovered_state <- as.character(recovered_origin[[c("Province.State")]])

if(identical(deaths_country, recovered_country) == identical(recovered_country, confirmed_country))
  {
  print("same country")
  }else
    {
      print("not same country")
    }

if(identical(deaths_state, recovered_state) == identical(recovered_state, confirmed_state))
{
  print("same state")
}else
{
  print("not same state")
}

```

All three data sources from the GHU and PDL show Hubei, China, as the origin for Covid-19.  The number of recovered, deaths, and confirmed cases were highest when this recording began on January 22nd, 2020.  Although it is possible that Hubei had the most awareness (through testing and recording) our agencies believe that the sustained increase of all three case counts suggest it was in fact close to the region of Hubei, China.

### Objective 2
```{r ob2}

i <- 0
column_num <- ncol(confirmed)
column_num_values <- confirmed[column_num - i]
column_num_b4 <-confirmed[column_num - i - 1]
column_values_sum <- sum(column_num_values == 0)
column_b4_sum <- sum(column_num_b4 == 0)

for (i in 1:column_num)
{
  if(column_values_sum != column_b4_sum)
  {
    break
  }
  i <- i + 1
  column_num_values <- confirmed[column_num - i]
  column_num_b4 <-confirmed[column_num - i - 1]
  column_values_sum <- sum(column_num_values == 0)
  column_b4_sum <- sum(column_num_b4 == 0)
  column_values_sum == column_b4_sum
}
print(i)

zero_values <- confirmed[confirmed$X1.20.21 == 0, ]
recent_case <- zero_values[zero_values$X1.21.21 != 0, c("Province.State", "Country.Region")]
recent_case

```

The most recent case occurred 79 days before the last day entered into the dataset.  The case occurred on January 21st, 2021, in Micronesia.  There is no Province/State associated to our database for Micronesia.  Micronesia was the only location that went from 0 cases to 1 or more cases.

### Objective 3
```{r ob3}

locations <- confirmed[c(72, 183), c(4:1)]
Hubei <- c(locations[1,1], locations[1,2])
Micronesia <- c(locations[2,1], locations[2,2])

dist_between <- round(distm(Hubei, Micronesia)*0.000621371, digits = 2)

paste(locations[2,3], " is ", dist_between, " miles away from ", locations[1,4], ", ", locations[1,3], ".", sep = "")

```

The origin of Covid-19 is suspected to be from Hubei, China.  The most recent confirmed case we can see is from Micronesia.  Micronesia is 2955.32 miles away from Hubei, China.

### Objective 4

#### Objective 4.1
```{r ob4.1}

column_num_deaths <- ncol(deaths)
total_deaths <- subset(deaths, select = c(1, 2, column_num_deaths))
total_deaths$StateRegion <- do.call(paste0, total_deaths[1:2])
colnames(total_deaths) = c("Province.State", "Country.Region", "Total.Deaths", "StateRegion")

column_num_recovered <- ncol(recovered)
total_recovered <- subset(recovered, select = c(1, 2,column_num_recovered))
total_recovered$StateRegion <- do.call(paste0, total_recovered[1:2])
colnames(total_recovered) = c("Province.State", "Country.Region", "Total.Recovered", "StateRegion")

column_num_confirmed <- ncol(confirmed)
total_confirmed <- subset(confirmed, select = c(1, 2,column_num_confirmed))
total_confirmed$StateRegion <- do.call(paste0, total_confirmed[1:2])
colnames(total_confirmed) = c("Province.State", "Country.Region", "Total.Confirmed", "StateRegion")

risk_score <- merge(total_deaths, total_recovered, by = "StateRegion")
risk_score <- subset(risk_score, select = c(1:4, 7))
risk_score <- merge(risk_score, total_confirmed, by = "StateRegion")
risk_score <- subset(risk_score, select = c(2:5, 8))
risk_score$Risk.Score <- risk_score$Total.Deaths / risk_score$Total.Recovered
risk_score$Death.Burden <- risk_score$Risk.Score * risk_score$Total.Confirmed

Global_Risk_Score <- sum(risk_score$Total.Deaths) / sum(risk_score$Total.Recovered)

Global_Risk_Score

head(risk_score[order(-risk_score$Risk.Score),], n = 12)
tail(risk_score[order(-risk_score$Risk.Score),], n = 22)

```

There are 22 rows under the way the data has been defined as Province/State or Country/Region areas that have 0 risk score.  They are Anguila (United Kingdom), Dominica, Falkland Islands (Malvinas, United Kingdom), Greenland (Denmark), Holy See, Jiangsu (China), Laos, Macau (China), Marshall Islands, Micronesia, New Caledonia (France), Ningxia (China), Northern Territory (Australia), Qinghai (China), Saint Helena (Ascension and Tristan da Cunha, United Kindgdom), Saint Kitts and Nevis, Saint Pierre and Miquelon (France), Samoa, Shanxi (China), Solomon Islands, Tibet (China), and Vanuatu.

The highest risk area is slightly harder to accurately define because seven areas have 0 recovered individuals reported and this makes the calculation infinite (and invalid).  Most of the areas have a high number of deaths except for New South Wales (Australia).  These areas are Belgium, Netherlands, New South Wales (Australia), Serbia, Sweden, United Kingdom, and US.  The areas with the highest risk scores that are not infinite are Martinique (France), Yemen, Spain, France, and MS Zaandam.  However, two of these areas have death and recovered counts below 100 which for me highlights problems with this parameter.  These high risk areas have risk scores from 0.286-0.602.  This is much higher (up to 15 times) than the global risk score, 0.038.  However, because there are some countries with no reported number of recovered individuals the global risk value is slightly inflated because the global number of deaths from those countries was included.  There is not a clear trend between risk score and burden score.  The populations have not been standardized and so the resulting burden score can vary greatly.

One problem with this dataset is that some countries have stopped reporting or have never reported the number of individuals recovered.  Sweden and the Netherlands never show a reported recovered case in the dataset.  The United States, for example, stopped reporting numbers on December 14th, 2020.

#### Objective 4.2
```{r ob4.2}

total_deaths2 <- subset(deaths, select = c(2,column_num_deaths))
total_deaths2$Sum.total <- rowSums(total_deaths2[-1])
total_deaths2 <- aggregate(x = total_deaths2$Sum.total, by = list(total_deaths2$Country.Region), FUN = sum)
colnames(total_deaths2) = c("Country.Region", "Total.Deaths")

total_recovered2 <- subset(recovered, select = c(2,column_num_recovered))
total_recovered2$Sum.total <- rowSums(total_recovered2[-1])
total_recovered2 <- aggregate(x = total_recovered2$Sum.total, by = list(total_recovered2$Country.Region), FUN = sum)
colnames(total_recovered2) = c("Country.Region", "Total.Recovered")

total_confirmed2 <- subset(confirmed, select = c(2,column_num_confirmed))
total_confirmed2$Sum.total <- rowSums(total_confirmed2[-1])
total_confirmed2 <- aggregate(x = total_confirmed2$Sum.total, by = list(total_confirmed2$Country.Region), FUN = sum)
colnames(total_confirmed2) = c("Country.Region", "Total.Confirmed")

risk_score2 <- merge(total_deaths2, total_recovered2, by = "Country.Region")
risk_score2 <- merge(risk_score2, total_confirmed2, by = "Country.Region")
risk_score2$Risk.Score <- risk_score2$Total.Deaths / risk_score2$Total.Recovered
risk_score2$Death.Burden <- risk_score2$Risk.Score * risk_score2$Total.Confirmed

Global_Risk_Score2 <- sum(risk_score2$Total.Deaths) / sum(risk_score2$Total.Recovered)
Global_Risk_Score2

kable(head(total_confirmed2[order(-total_confirmed2$Total.Confirmed),], n = 5))


kable(head(total_recovered2[order(-risk_score2$Total.Recovered),], n = 5))


kable(head(total_deaths2[order(-risk_score2$Total.Deaths),], n = 5))

```

The top five countries with the most confirmed cases are the US, Brazil, India, France, and Russia.  The top five countries with most recovered are India, Brazil, Russia, Turkey, and Italy.  This list will not include the countries that never reported or stopped reporting recovered cases.  The top five countries with the most deaths are the US, Brazil, Mexico, India, and the United Kingdom.

### GitHub Log
```{bash gitlog} 
git log --pretty=format:"%nSubject: %s%nAuthor: %aN%nDate: %aD%nBody: %b"
```





